\normalfont \normalsize \textsc{Northwestern University} \\ [25pt] \horrule{0.5pt} \\[0.4cm] \huge Masters Thesis \\ \horrule{2pt} \\[0.5cm]

### 2016-08-19 Gradschool/thesis/msthesis.tex
expanded cuda section

@@

\subsection{GPU Implementation}
Vectorization of the code produces a performance improvement due to SIMD instruction sets available on many modern microprocessor architectures. This allows for some on-chip parallelism that the compiler may use to unroll loops within the machine code and increase efficiency. Beyond this however the process of recasting the computation into clear matrix and tensor operations illuminates a higher level of parallelism possible on the hardware level. A substantial amount of both academc and industrial research and development has centered around efficient parallel implementations of the equivalent operations, with impressive results.

>>>>

$h_{\mu \nu}$ is a two index object in a four dimensional spacetime, which in the general case would present with sixteen independent components each of which would need to be explicitly determined to form a complete picture of the physics. However in our specific case it will turn out that there are far fewer real independent components then this, which can be observed by chiseling away at the above expression in the following manner. Firstly, the metric of general relativity is taken to be symmetric, because if we assume that the tensor $dx^{\mu}dx^{\nu}$ is symmetric (i.e. $dx^{\nu}$ and $dx^{\nu}$ commute) then only the symmetric part of any $g_{\mu \nu}$ contributes when the quantity $g_{\mu \nu}dx^{\mu}dx^{\nu}$ is computed. Thus we have

\begin{align}
h_{\mu \nu} = h_{\nu \mu}
\end{align}

Which, for a sixteen component tensor, makes six of the components redundant. This leaves us with ten. Furthermore, there is \textit{gauge freedom} present in the linearized vacuum Einstein equation. This is best illustrated with reference to the equations of classical electrodynamics, which describe the motion of charged particles subject to the electric and magnetic vector potentials, $V$ and $\vec{A}$ respectively. It is somewhat easily demonstrable that the same equations of motion for electrically charged particles are unchanged when $V$ and $\vec{A}$ are subject to transformations of the form

\begin{align}
\vec{A} &\rightarrow \vec{A} + \nabla \Psi \\
V &\rightarrow V -\frac{\partial \Psi}{\partial t}
\end{align}

The function $Psi$ is known as the \textit{gauge function} and it can be absolutely anything we want, as long as we adjust the potentials accordingly. Within such a framework, one can impose certain conditions on the potentials that simplify the mathematics, such as requiring that

\begin{align}
\nabla \cdot \vec{A} + \frac{1}{c}\frac{\partial V}{\partial t} &= 0 \\
\rightarrow \partial^{\mu}A_{\mu} &= 0
\end{align}

It is somewhat counterintuitive why imposing this particular condition on $\vec{A}$ and $V$ themselves is equivalent to transforming the potentials via the gauge function, but indeed it is, it is just hidden! While we have made no reference to $\Psi$, we have \textit{implicitly} assumed that there is such a $\Psi$ that will \textit{make} the restriction on $\vec{A}$ and $V$ true. It actually is there inside the $V$ and $\vec{A}$ written in equation \textbf{EQUATION}. We do not have to state \textit{what it is}, as long as we know that it exists. A proof that there legitimately is such a gauge function exists but is out of the scope of this outline. 
There is a one-to-one correspondance of this formalism with the equivalent gauge in general relativity, which is known as the Lorentz gauge. In the same sense as equation \textbf{EQUATION}, it turns out that the equations of motion for massive particles in a slightly perturbed spacetime are invariant under coordinate transformations of the form

\begin{align}
x^{\mu} \rightarrow x'^{\mu} - \xi^{\mu}
\end{align}

In particular, we can show that there exists a coordinate transformation such that

\begin{align}
\partial_{\mu}V_{\nu} = \partial_{\mu}\partial_{\alpha}h^{\alpha}_{\nu} - \frac{1}{2}\partial_{\mu}\partial_{\nu}h = 0
\end{align}

\textbf{PUT PROOF HERE}

\begin{align}
\text{PUT PROOF HERE}
\end{align}

Thus coordinate transformations that obey $\Box \xi = 0$ will allow us to impose the Lorentz gauge. Within this gauge, the linearized, vacuum Einstein equation becomes simply

\begin{align}
\Box h_{\mu \nu} = 0
\end{align}

This is a four dimensional wave equation for the remaining components of $h_{\mu \nu}$. Moreover, we have shown that under coordinate transformations that do not change the underlying physics, there exists an explicit relationship between components of the metric given by equation \textbf{EQUATION}. In the same way that symmetry shows that six of the metric components to be functions of the others, The index $\nu$ indicates that in four dimensions there are four additional explicit relationships between metric components that correspond to a loss of four additional degrees of freedom. This leaves us with six.

\section{Bayesian Statistics}
We select a Bayesian posterior probability as our main analysis technique as its ability to quantify and propagate uncertainties in parameter estimates is well documented by \textbf{CITATION} and \textbf{PROBABLY OTHER CITATIONS}. Bayes' Theorem follows from the symmetric definition of conditional probability depicted visually in figure \textbf{FIGURE}. Let the joint probability evaluated at any point $(x,y)$ in the plane be denoted $P(x,y)$. We may take a cross section through the joint probability distribution along any direction we choose, let us call such a cross section parallel to the $x$ axis the conditional probability of $x$ given $y$, and denote it $P(x|y)$. A perpandicular cross section would simply be $P(y|x)$. We see that we can compute $P(x,y)$ in two equivalent ways: scaling $P(x|y)$ by $P(y)$ or by scaling $P(y|x)$ by $P(x)$. Thus we have that

\begin{align}
P(x|y)P(y) &= P(y|x)P(x) \\
P(x|y) &= \frac{P(y|x)P(x)}{P(y)}
\end{align}

Which is Bayes' theorem. The denominator of the expression on the right hand side is often rewritten as a sum over all possible situations in which $y$ is an outcome. Reading $\neg x$ as "not x", we have

\begin{align}
P(y) &= P(y|x)P(x) + P(y|\neg x)P(\neg x) \\
&= \sum_{k=1}^K P(x|y_k)P(y_k) 
\end{align}

 Assuming a signal is present, the measured strain in a gravitational wave interferometer $d(t)$ can be decomposed into time dependent strain and noise components

\begin{align}
d(t) = h(t) + n(t)
\end{align}

then the posterior probability of a gravitational wave signal with parameters $\mu$ is given by

\begin{align}
P[\vec{\mu}|d(t)] = \frac{P[d(t)|\vec{\mu}]P(\vec{\mu})}{P[d(t)]}
\end{align}

Extending equation \textbf{EQUATION} to the case of a continuous state space we have for the denominator

\begin{align}
P[d(t)] &= \int_{\Omega}P[d(t)|\vec{\mu}]P(\vec{\mu})d\vec{\mu}
\end{align}

Where $\Omega$ indicates integration over the entire probabilistic state space.

\subsection{Harmonic Mode Decomposition}

For a single detector $k$ We can write our gravitational wave signal $h_k(t) = h_{+,k}(t) - ih_{\times, k}(t)$ as a sum over products of harmonic mode strain values $h_{l,m}(\vec{\lambda})$ and corresponding spin-weighted spherical harmonics, scaled by the reference distance and antenna factor:

\begin{align}
h_k(t) = \frac{D_r}{D}F_k(\theta)\sum_{(l,m)}h_{k, (l,m)}(t, \vec{\lambda})Y_{(l,m)}(\vec{\theta})
\end{align}

Where $D_r$ is the luminosity reference distance to the binary. We substitute this expression for the signal into the expression for the likelihood ratio we find, taking the logarithm, that

\begin{align}
\mathcal{L} &= \prod_k e^{-\frac{1}{2}\braket{d - h_k|d-h_k} + \braket{d|d}} \\
\ln\mathcal{L} &= -\frac{1}{2}\sum_{k}\left[\braket{d - h_k|d-h_k} + \braket{d|d}\right] \\
&= \frac{1}{2}\sum_{k}\left[-\cancel{\braket{d|d}}+\braket{d|h_k}+\braket{h_k|d}+\braket{h_k|h_k} + \cancel{\braket{d|d}}\right] \\
\end{align}

Now, since for complex vector spaces $\braket{a|b} = \braket{b|a}*$, we have that $\braket{d|h_k} + \braket{h_k|d} = 2\Re\braket{h_k|d}$. Then,

\begin{align}
\ln\mathcal{L} &= \frac{1}{2}\sum_{k}\left[2\braket{h_k|d} + \braket{h_k|h_k}\right] \\
&= \frac{1}{2} \sum_{k}\Re\left[\braket{2\frac{D_r}{D}F_k(\vec{\theta})\sum_{(l,m)}h_{k,(l,m)}(\lambda)Y_{(l,m)}(\vec{\theta})|d}\right] \\
&+ \left[\frac{D_r}{D}\right]^2F_k^2(\vec{\theta})\sum_{(l,m),(l',m')}h_{(l,m)}(\vec{\lambda})h_{(l',m')}(\vec{\lambda})Y_{(l,m)}(\vec{\theta})Y_{(l',m')}(\vec{\theta})
\end{align}

The defining feature of this form of the likelihood is the separation between intrinsic and extrinsic parameters. The intrinsic parameters of the binary exist only within the $h_{k,(l,m)}$, whereas the parameters that depend on the observer are contained entirely within the antenna factor and spherical harmonics. Work by \textbf{CITE CHRIS AND RICHARD} has shown that this can be exploited to accelerate computation of the likelihood by precomputing the orbital dynamics of the binary and marginalizing out all of the extrinsic parameters with a Monte Carlo integration.

@@

\subsection{GPU Implementation}
Vectorization of the code produces a performance improvement due to SIMD instruction sets available on many modern microprocessor architectures. This allows for some on-chip parallelism that the compiler may use to unroll loops within the machine code and increase efficiency. Beyond this however the process of recasting the computation into clear matrix and tensor operations illuminates a higher level of parallelism possible on the hardware level. A substantial amount of both academc and industrial research and development has centered around efficient parallel implementations of the equivalent operations, with impressive results. Reaching the limit of the sample throughput with conventional serial python libraries, our investigation turned to GPU based acceleration as the only means forward. The following is the result of this investigation.

\subsection{Hardware Limitations and Logical Programming}
GPUs work significantly differently from CPUs and must be paired with one to operate, they serve only as accelerators for complex operations and not as standalone units. Data must be passed back and forth from host memory on the node to RAM placed directly on the card itself through expensive transfers via the PCI bus. This is known as device global memory. While the onboard RAM on bleeding edge cards exceeds 12GB, memory transfers are the dominant cost of many GPU programs and must be minimized. Once the data exists on the device, it is accessed and manipulated by anywhere between nine and fifteen logical cores called Streaming Multiprocessors. These cores operate concurrently and contain a smaller granularity of parallel processing units that \textit{may} operate concurrently depending on the individual resources required by the requested operations.
Programming in CUDA involves launching groups of threads called \textit{grids} that are further subdivided into \textit{blocks}. The blocks themselves are indexed by local variables that are rapidly accessible by all the contained threads, and the threads themselves all have access to indices placed within registers that exist for the lifetime of the thread in question. Both the blocks and and grids can be up to three dimensional and thus each block and thread may have access to up to three registers containing block indices and three containing thread indices. In addition thread blocks have access to an extremely high speed shared memory space where ideally most of the computations are performed, however it is limited in size to \textbf{SIZE} and is not designed to hold the entire target dataset. One major design goal for CUDA programs is implementing a process which copies portions of memory from the global space to the shared, processes it, and copies it back, while using the thread and block indices to blanket the dataset with threads and control which threads access what data. The current CUDA API allows for the logical launch of \textbf{HOW MANY} blocks with a maximum of 1024 threads per block, for a total of \textit{HOW MANY} possible "concurrent" threads. It is important to note that these threads are not \textit{physically} concurrent. Indeed, if each of the logical threads containing six integers in registers were to exist simultaneously, the GPU would be storing \textbf{HOW MANY} registers at once! Similarly perposterous amounts of shared memory would have to exist. In actuality, the shared memory and registers are located on the logical cores themselves, and blocks thread blocks line up to be processed. How many are processed at once is heavily dependent on the resources requested by the blocks themselves. Defining too many local variables in a kernel function can lead to \textit{register pressure} and decreased \textit{occupancy} of the multiprocessors on the GPU, a similar effect can come from requesting too much shared memory. Furthermore, the blocks are not guaranteed to execute in any particular order. While the compiler is responsible for sorting out how many blocks can be processed at a time as well as the order, the programmer is responsible for performance tuning with regard to optimizing block occupancy. This is often a late stage design consideration, but also leads to some of the greatest returns after algorithmic optimization has been completed.

\subsection{PyCuda} 
PyCuda is a flexible library for GPU scripting that includes a broad spectrum of GPU based array operations built to mimic the usage of NumPy arrays within a program. It allows for relatively easy access to the Nvidia CUDA API directly from within Python, and many of its operations are directly compatible with NumPy arrays with careful casting to the correct data types. Certainly one of the most challenging aspects of incorporating GPU code into a higher level language such as Python is the interface layer. As it stands much of the widely used LAL software consists of C code bound to Python through complex Swig interfaces, and while this is a perfectly valid way to access faster libraries, it can often be deceptively difficult to work with as many of the resulting Python objects contain read-only data structures masquerading as dynamic Python objects. This sometimes yields unpredictable behavior and it was decided this should be avoided as a way to interface with CUDA C. PyCuda is a very elegant solution to this problem as it seamlessly bridges this gap using the native GPUArray class, and a set of functions that create GPUArray copies of NumPy arrays that are built initially on the node CPU hosts. 
With a smaller group of developers, the built-in GPUArray class contained within PyCuda does not yet support \textit{all} of the rich spectrum of high level functions available within the NDArray NumPy class. Perhaps more importantly however, PyCuda allows for custom CUDA kernel functions to be written as pure CUDA C directly within special containers called source modules, that are contained within the main body of the Python code. PyCuda actually calls the real Nvidia compiler at runtime and links the resulting temporary executable to the main program. One may call these functions with NumPy arrays as arguments, as if they were normal Python functions. This is perhaps the most important feature of PyCuda as it allows the developer to tune the customization of the code to whatever level is desired. Those intermediate operations that may be handled with the built in methods of the GPUArray are performed as such. More complex operations that are not supported can be custom built with source modules. If desired, the entire program may consist of CUDA C with no more then basic startup routines written in actual Python meant to instantiate the original data and pass it down to the GPU. This particular work used mostly custom kernels, for the simple reason that applying the same matrix operation to a contiguous block of memory containing multiple slightly different copies constitutes tensor operations similar in nature to the results of section \textbf{SECTION}, for which there is no general support for in PyCuda.
